Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-22 paper-filmat.tex
Version as just handed in

\begin{center}
  \begin{minipage}{0.6\textwidth}
  \emph{To the anonymous referees:}
  \bigskip

  Thank you for your volunteer work. It is very much appreciated. I am looking
  forward to your comments.

  \hfill -- the anonymous author
  \end{minipage}
\end{center}
\bigskip

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{Olivia Caramello's research
program~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories.}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

@@

\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
multiverse view in set theory~\ref{hamkins:multiverse}. In fact, the
topos-theoretic multiverse can be regarded as an extension of the set-theoretic
multiverse: While Hamkins proposes to embrace all models of set theory (not
necessarily all of them not equally), we propose to embrace all toposes (again
not necessarily all of them not equally). As every model~$M$ of set theory
gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
topos-theoretic one.

@@

One can always mechanically eliminate the topos machinery from such a proof, by
translating all intermediate statements following the Kripke--Joyal translation
rules and unwinding the constructive soundness proof of
Theorem~\ref{thm:reasoning}. This unwinding typically turns transparent
internal proofs into complex external proofs -- proofs which one may not have
found without the problem-adapted internal language provided by a
custom-tailored topos.

@@

* XXX phone call analogy
FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
* XXX A topos is more than its semantics
FUTURE: Check references everywhere (give good pointers)

>>>>

\begin{center}
  \begin{tikzpicture}
    \node (comment) at (0,0) {
      \begin{minipage}{0.6\textwidth}
      \emph{To the anonymous referees:}
      \bigskip

      Thank you for your volunteer work. It is very much appreciated. I am looking
      forward to your comments.
      \bigskip

      \hfill -- the anonymous author
      \end{minipage}
    };

    \node[draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt] [fit = (comment)] (foo) {};
  \end{tikzpicture}
\end{center}
\bigskip

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and that we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{Olivia Caramello's research
program~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories. \textbf{\textcolor{red}{NOTE TO EDITOR: Insert cross reference to Olivia's chapter of
this book.}}}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

% FUTURE:
% \begin{rem}It might be helpful to picture mathematicians in a topos~$\E$. For
% exchanging ideas, they use the same mathematical language like we do, but their
% mathematical results seem unfamiliar or even wrong to us. Only when we remember
% to switch on the Kripke--Joyal translation engine their claims start making
% sense.\end{rem}

@@

\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
multiverse view in set theory~\cite{hamkins:multiverse}. In fact, the
topos-theoretic multiverse can be regarded as an extension of the set-theoretic
multiverse: While Hamkins proposes to embrace all models of set theory (not
necessarily all of them not equally), we propose to embrace all toposes (again
not necessarily all of them not equally). As every model~$M$ of set theory
gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
topos-theoretic one.

@@

One can always mechanically eliminate the topos machinery from such a proof, by
translating all intermediate statements following the Kripke--Joyal translation
rules and unwinding the constructive soundness proof of
Theorem~\ref{thm:reasoning}. This unwinding typically turns transparent
internal proofs into complex external proofs -- proofs which one might not have
found without the problem-adapted internal language provided by a
custom-tailored topos.

@@

FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
FUTURE: A topos is more than its semantics
FUTURE: Check references everywhere (give good pointers)
